Instructions for developing new pages for the MarineOmics webpage

### 2023-12-13 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd

The purpose of this tutorial is to provide an example of identifying loci- and gene-level methylation in bisulfite-treated DNA sequence data. The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data. When aligning to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines. For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated.

For a good review of epigenetic mechanisms including DNA methylation in marine organisms see @doi:10.1146/annurev-marine-010318-095114. @https://doi.org/10.1111/1755-0998.13542 offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS). The tutorial below is based on WGBS, though the general workflow would be consistent.

@@

>>>>

Thanks to advances in sample preparation and sequencing methods, there is an influx of research examining epigenomics in non-model systems. Broadly, epigenetics can be defined as changes to gene expression that do not arise from changes in the DNA sequence. DNA methylation, or addition of a methyl (CH<sub>3</sub>) group to a cytosine base adjacent to a guanine (CpG) is one of the more commonly studies epigenetic mechanisms, partly because of its environmental sensitivity and potential role in phenotypic plasticity. For a good review of epigenetic mechanisms including DNA methylation in marine organisms see @doi:10.1146/annurev-marine-010318-095114.

The purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data and discuss needs for potential downstream applications. The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data. @https://doi.org/10.1111/1755-0998.13542 offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS). When aligning bisulfite-converted data to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines. For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated. Since bisulfite treatment can damage DNA, newer library preparation and sequencing methods (ex. [NEBNext® Enzymatic Methyl-seq Kit; EM-seq](https://www.neb.com/en-us/products/e7120-nebnext-enzymatic-methyl-seq-kit#Protocols,%20Manuals%20&%20Usage) are starting to gain popularity. As EM-Seq produces loci-level data, alignment of EM-Seq data is similar to WGBS.

The tutorial will cover the following:

- Sequence quality assessment
- Read alignment
- Methylation quantification
- File conversions
- Potential downstream applications. Although this tutorial will not provide code for downstream applications, it will link to other open-access resources and examples.

The tutorial below is based on WGBS from a species of marine intertebrate (*Montipora capitata* coral) examined in @https://doi.org/10.1111/1755-0998.13542, and all raw data can be accessed under NCBI Bioproject PRJNA691891 if you would like to follow the tutorial with this dataset. The general workflow would be consistent for other library preparation and sequencing methods that produce loci-level data. The tutorial indicates places where special analytical consideration should be given to other data types.

@@

# Potential downstream applications

- Gene-level methylation analysis: people do this, but worth considering the utility of converting loci-level resolution data to gene-level resolution data
- methylKit
- IGV
- Analysis with other datasets (ex. pop gen, RNA-Seq)
